The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-29 paper.tex
Texto do Adriano

>>>>

Terra is a system that implements a specif programming model for wireless 
sensor networks~\cite{Branco2015a}.
Terra is based on virtual machine concepts that uses an implementation of Céu 
language as its scripting language.

Terra programming model combines the use of configurable component-based 
virtual machines with a reactive scripting language which can be statically 
analyzed to avoid unbounded execution and memory conflicts.
This approach allows the flexibility of remotely uploading code on motes to be 
combined with a set of guarantees for the programmer.
The choice of the specific set of  components in a virtual machine 
configuration defines the abstraction level seen by the application script.

Terra uses \ceut as its scripting language and implements a component-based 
virtual machine \vmt to be customized for different application domains.
\ceut implements a variation of Céu programming language.
This variation, on one hand, excludes some of Céu's characteristics and, on the 
other hand, includes some language extensions.
We use Céu as Terra's scripting language because of its reactive nature and its 
high-level control primitives and compile-time safety guarantees.
We want only the VM components to escape the safety analysis, so we took out 
the facility to include arbitrary C code, but we did maintain all of Céu's 
original control structures.

The Terra Virtual machine \vmt, with its customizations, runs on WSN nodes with 
limited resources like MicaZ\cite{Memsic2009b} and TelosB\cite{Memsic2009} 
motes.
We built \vmt using the nesC programming language~\cite{Gay2003} and the TinyOS 
operating system~\cite{Levis2004}.
We used TinyOS, basically, because it is the operating system for WSN  most 
ported to different mote types.

Figure \ref{fig:system} presents the three basic elements of Terra -- a 
scripting language (\ceut), a set of customized pre-built components, and an 
embedded virtual-machine engine.
The \vmt runtime environment is the virtual machine combined with a set of 
customized components.

\begin{figure}[!htb]
 \centering
 \includegraphics[width=0.7\columnwidth]{fig_system}
\caption{Terra programming system basic elements.}
 \label{fig:system}
\end{figure}

A \ceut program is compiled to a bytecode file that can then be disseminated to 
the network nodes, where it is interpreted by the \vmt, which implements the 
bytecode interpreter, the execution model, the code dissemination service, and 
some specific customized components.
Figure \ref{fig.impl-t} illustrates the Terra compilation process.

\begin{figure}[!htb]
\centering
\includegraphics[width=1.0\columnwidth]{impl-t}
\caption{Terra Compilation process: from the source code in \ceut to the final 
bytecode.
\label{fig.impl-t}
}
\end{figure}

Similar to the Céu compiler process, the Terra process also has the three main phases:
Parsing, Temporal Analysis, Code Generation.
The Terra parsing phase uses the original Céu parser adding some extensions to 
parse specific \ceut elements as the data types, the data structures, and the 
configuration block.
A configuration block defines all events, functions, and data structures 
available in a specific \vmt customization.
The Terra Temporal Analysis phase  only differs in the code output. Instead of 
generating code in C, Terra generates the \vmt assembly code using the original 
Céu labels for trails entry points.
In the final phase for code generation, Terra compiler generates the \vmt 
bytecode file using the program address in place of Céu labels.
The use of address allows the Céu scheduler to execute directly a trail without 
a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all Céu controls and variables.
All Céu data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original Céu run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
